---
title: "CRC Simulation Analysis (Generated Data)"
format:
  html:
    toc: true
execute:
  warning: false
  message: false
---

## Overview

This notebook runs the full generated-data CRC workflow:

1. Create synthetic contingency-table data.
2. Extract uncaptured ground truth (all capture columns are 0).
3. Apply manipulations (suppression).
4. Run one or many CRC models with one or many configurations.
5. Compare model intercept estimates to ground truth.

## Setup

```{r}
library(crcsim)
library(dplyr)
```

## Parameters

```{r}
seed <- 2024
# n_individuals <- 300000
n_individuals <- 500
p_captures <- c(0.45, 0.35, 0.25, 0.20)
p_strata <- c(0.60, 0.40)

methods <- c("poisson", "negbin")
config <- list(
    f0_05 = list(direction = "forward", threshold = 0.05),
    b0_05 = list(direction = "backward", threshold = 0.05),
    fb0_05 = list(direction = "both", threshold = 0.05)
)

suppress <- 10
```

## Single-Run Analysis

```{r}
single_run <- run_crc(
    mode = "single",
    p_captures = p_captures,
    p_strata = p_strata,
    n_individuals = n_individuals,
    methods = methods,
    config = config,
    suppress = suppress,
    formula_selection = "stepwise",
    seed = seed
)

single_run
```

## Single-Run TMLE Analysis

This is not working, I'll need to dig in here next.

```{r}
single_tmle_run <- run_crc(
    mode = "single",
    p_captures = p_captures,
    p_strata = p_strata,
    n_individuals = n_individuals,
    methods = methods,
    config = config,
    suppress = suppress,
    formula_selection = "tmle",
    opts_tmle = list(
        list_pair = c("capture_1", "capture_2"),
        funcname = "logit",
        nfolds = 2,
        margin = 0.005,
        estimator = "TMLE",
        expansion_mode = "exact",
        sample_size = 100000,
        warn_expanded_rows = 1000000
    ),
    seed = seed
)

single_tmle_run
```

## Bootstrap Analysis

```{r}
n_bootstraps <- 5

bootstrap_run <- run_crc(
    mode = "bootstrap",
    n_bootstraps = n_bootstraps,
    p_captures = p_captures,
    p_strata = p_strata,
    n_individuals = n_individuals,
    methods = methods,
    config = config,
    suppress = suppress,
    formula_selection = "stepwise",
    seed = seed
)

bootstrap_run
```

## Summary

```{r}
bootstrap_summary <- bootstrap_run |>
    group_by(Method, Model, group) |>
    summarize(
        mean_estimate = mean(estimate, na.rm = TRUE),
        mean_ground = mean(ground, na.rm = TRUE),
        mean_bias = mean(estimate - ground, na.rm = TRUE),
        mean_time_to_convergence_sec = mean(
            time_to_convergence_sec,
            na.rm = TRUE
        ),
        median_time_to_convergence_sec = median(
            time_to_convergence_sec,
            na.rm = TRUE
        ),
        .groups = "drop"
    )

bootstrap_summary
```

## Convergence Timing Benchmark

```{r}
convergence_benchmark <- bootstrap_run |>
    group_by(Method, Model) |>
    summarize(
        n_runs = dplyr::n_distinct(Run),
        mean_time_to_convergence_sec = mean(
            time_to_convergence_sec,
            na.rm = TRUE
        ),
        p95_time_to_convergence_sec = stats::quantile(
            time_to_convergence_sec,
            probs = 0.95,
            na.rm = TRUE
        ),
        .groups = "drop"
    ) |>
    arrange(mean_time_to_convergence_sec)

convergence_benchmark
```

## Convergence Time Boxplot

```{r}
library(ggplot2)

bootstrap_run |>
    ggplot(aes(x = Model, y = time_to_convergence_sec, fill = Method)) +
    geom_boxplot(alpha = 0.7) +
    facet_wrap(~Method, scales = "free_x") +
    labs(
        title = "Distribution of Convergence Times by Method and Model",
        x = "Model Configuration",
        y = "Time to Convergence (seconds)",
        fill = "Method"
    ) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, face = "bold")
    )
```
