---
title: "Simulation Data"
author:
    - name: "Matthew Carroll"
---

```{r setup, include=FALSE}
devtools::load_all("..", quiet = TRUE)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2, warn.conflicts = FALSE)

theme_set(theme_minimal(base_size = 12))
```

# Purpose

This notebook is the concurrent analysis for the structured simulation
constructor. It checks that the complete population, observed records, hidden
all-zero records, and truth metadata remain logically consistent while the
simulation scenarios are developed.

## Data Contract and Scopes

Each simulation has three set splits. `full` contains every
generated individual, including people missed by every capture list.
`observed` is the union of the capture lists and therefore excludes the
all-zero history. `unobserved` contains exactly those all-zero histories. 
It is outlined in the equation below. The summary object keeps these views labeled as `full`, `observed`, and
`unobserved`.

$$
n(\text{full}) = n(\text{observed}) + n(\text{unobserved})
$$

The simulation also retains generating truth, such as fixed probabilities or
individual capture probabilities, so diagnostics can compare what was generated
with what was visible.

# Simulation Scenarios

The scenarios can be split into several different top-level ideas:

1. Baseline
2. Covariate-Dependent
3. Latent Heterogeneity

## Baseline Family

The first scenario builds a baseline family of fixed-probability
conditional-independence captures. The 3 components of the baseline family are
differentiated by the capture probabilities:

1. Mid Capture: Probabilities are 0.50.
2. Low Capture: Probabilities are 0.20.
3. Imbalanced Capture: Probabilities are 0.80, 0.35, and 0.10.

```{r scenarios}
baseline_scenarios <- list(
    mid_capture = c(0.50, 0.50, 0.50),
    low_capture = c(0.20, 0.20, 0.20),
    imbalanced_capture = c(0.80, 0.35, 0.10)
)

# Loads the baseline data into the scenario_data list
scenario_data <- lapply(names(baseline_scenarios), function(scenario_name) {
    simulation <- simulate_data(
        n_individuals = 5000,
        n_captures = 3,
        p_captures = baseline_scenarios[[scenario_name]],
        seed = 20261005
    )
    simulation$metadata$scenario_name <- scenario_name
    return(simulation)
})
names(scenario_data) <- names(baseline_scenarios)
```

## Covariate-Dependent Captures

The package helper currently applies one probability vector to every list. For
this notebook scenario, each list receives its own logistic model while the
underlying conditional-independence structure remains explicit:

```{r covariate-scenario}
covariate_columns <- c("covariate_1", "covariate_2")
capture_columns <- paste0("capture_", seq_len(3))
covariate_intercepts <- c(-1.2, -0.8, -1.5)
covariate_coefficients <- rbind(
    capture_1 = c(1.0, -0.5),
    capture_2 = c(0.4, 0.8),
    capture_3 = c(-0.7, 1.2)
)

covariate_ranges <- data.frame(
    distribution = c("uniform", "uniform"),
    p1 = c(0, 0),
    p2 = c(1, 1),
    dtype = c("integer", "integer")
)

covariate_simulation <- simulate_data(
    n_individuals = 5000,
    covariate_ranges = covariate_ranges,
    alpha = covariate_intercepts,
    beta = covariate_coefficients,
    seed = 20261005
)

add_categorical_covariate <- function(data) {
    data$covariate_group <- factor(
        ifelse(data$covariate_1 == 0, "zero", "one"),
        levels = c("zero", "one")
    )
    return(data)
}

covariate_simulation$full_data <- add_categorical_covariate(
    covariate_simulation$full_data
)
covariate_simulation$observed_data <- add_categorical_covariate(
    covariate_simulation$observed_data
)
covariate_simulation$unobserved_data <- add_categorical_covariate(
    covariate_simulation$unobserved_data
)
covariate_simulation$metadata$scenario_name <- "covariate_dependent"
scenario_data[["covariate_dependent"]] <- covariate_simulation

latent_simulation <- simulate_data(
    n_individuals = 5000,
    n_captures = 3,
    p_captures = rep(0.5, 3),
    latent_sd = 1,
    seed = 20261005
)
latent_simulation$metadata$scenario_name <- "latent_heterogeneity"
scenario_data[["latent_heterogeneity"]] <- latent_simulation

pairwise_simulation <- simulate_data(
    n_individuals = 5000,
    n_captures = 3,
    p_captures = rep(0.5, 3),
    dependence_pair = c(1, 2),
    dependence_strength = 2,
    seed = 20261005
)
pairwise_simulation$metadata$scenario_name <- "direct_pairwise_dependence"
scenario_data[["direct_pairwise_dependence"]] <- pairwise_simulation

class_simulation <- simulate_data(
    n_individuals = 5000,
    latent_class_probabilities = c(0.75, 0.25),
    latent_class_p_captures = rbind(
        c(0.8, 0.8, 0.8),
        c(0.2, 0.2, 0.2)
    ),
    seed = 20261005
)
class_simulation$metadata$scenario_name <- "latent_classes"
scenario_data[["latent_classes"]] <- class_simulation

graph_simulation <- simulate_data(
    n_individuals = 5000,
    n_captures = 3,
    p_captures = rep(0.5, 3),
    dependence_edges = data.frame(
        from = c(1, 2),
        to = c(2, 3),
        strength = c(2, -2)
    ),
    seed = 20261005
)
graph_simulation$metadata$scenario_name <- "graph_dependence"
scenario_data[["graph_dependence"]] <- graph_simulation

covariate_data <- covariate_simulation$full_data[, covariate_columns]
covariate_probabilities <- covariate_simulation$truth$capture_probabilities
capture_columns <- covariate_simulation$metadata$capture_columns

scenario_summaries <- lapply(scenario_data, function(simulation) {
    summarize_simulation(
        simulation,
        capture_columns = simulation$metadata$capture_columns
    )
})
```

The coefficients are known because they generated the data. The following
plot checks that the lists have distinct covariate-dependent probabilities.

For list $j$, the conditional capture probability is
$\operatorname{logit}^{-1}(\alpha_j + x^T\beta_j)$. The lists remain
conditionally independent given the covariates, but their marginal capture
rates can differ because each list has its own coefficient row.

```{r covariate-probabilities}
probability_data <- cbind(
    covariate_data,
    as.data.frame(covariate_probabilities)
) |>
    tidyr::pivot_longer(
        cols = dplyr::all_of(capture_columns),
        names_to = "capture",
        values_to = "probability"
    )

ggplot(
    probability_data,
    aes(x = factor(covariate_1), y = probability, fill = capture)
) +
    geom_boxplot(position = position_dodge(width = 0.8)) +
    facet_wrap(~covariate_2) +
    labs(
        title = "True list-specific capture probabilities",
        x = "Covariate 1",
        y = "Capture probability",
        fill = "List"
    )
```

The reusable summary also records the corresponding full-population truth as
one distribution row per capture list.

```{r probability-summary}
probability_summary <- scenario_summaries$covariate_dependent$capture_probabilities |>
    filter(scope == "full")

ggplot(
    probability_summary,
    aes(x = capture, y = mean, fill = capture)
) +
    geom_col(width = 0.7, show.legend = FALSE) +
    geom_errorbar(
        aes(ymin = mean - sd, ymax = mean + sd),
        width = 0.15
    ) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
        title = "Mean true capture probability by list",
        x = "List",
        y = "Mean probability"
    )
```

The same reusable summary object provides categorical covariate proportions
without requiring a second data-preparation path.

```{r categorical-covariate-summary}
categorical_summary <- scenario_summaries$covariate_dependent$covariate_levels |>
    filter(
        covariate == "covariate_group",
        scope %in% c("full", "observed", "unobserved")
    )

ggplot(
    categorical_summary,
    aes(x = level, y = proportion, fill = scope)
) +
    geom_col(position = position_dodge(width = 0.8)) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
        title = "Categorical covariate proportions by simulation scope",
        x = "Covariate group",
        y = "Proportion",
        fill = "Scope"
    )
```

# Structural Validation

The all-zero histories are available only because this is a simulation. They are
retained as truth for evaluation and are excluded from the observed view.

```{r validation}
validation_results <- bind_rows(lapply(scenario_summaries, function(summary) {
    counts <- summary$counts
    full_count <- counts$n_individuals[counts$scope == "full"]
    observed_count <- counts$n_individuals[
        counts$scope == "observed"
    ]
    unobserved_count <- counts$n_individuals[
        counts$scope == "unobserved"
    ]

    stopifnot(
        observed_count + unobserved_count == full_count,
        all(
            summary$capture_patterns$proportion[
                summary$capture_patterns$scope == "full"
            ] >=
                0
        )
    )

    data.frame(
        scenario = summary$metadata$scenario_name,
        n_total = full_count,
        n_observed = observed_count,
        n_unobserved = unobserved_count,
        observed_fraction = observed_count / full_count
    )
}))

validation_results
```

The observed fraction is not a model estimate. It is a known simulation quantity
used to describe how much of the generated population is visible to the
capture-recapture estimator.

This distinction is important for later estimator comparisons: an estimator
will only see the observed union, while the simulation can evaluate its
population error against `n_total`. The hidden all-zero count is therefore an
evaluation target, not an input supplied to the estimator.

```{r validation-plot}
ggplot(
    validation_results,
    aes(x = scenario, y = observed_fraction, fill = scenario)
) +
    geom_col(width = 0.7, show.legend = FALSE) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
        title = "Observed fraction under simulation scenarios",
        x = NULL,
        y = "Observed fraction"
    )
```

# Capture-History Behavior

The next plot compares the frequency of nonzero capture histories. The all-zero
history is deliberately shown separately because it is the hidden population
component in an observed capture-recapture dataset.

```{r capture-histories}
history_data <- bind_rows(lapply(scenario_summaries, function(summary) {
    summary$capture_patterns |>
        filter(!is.na(history)) |>
        mutate(
            scenario = summary$metadata$scenario_name,
            observed = history !=
                paste(
                    rep("0", length(summary$metadata$capture_columns)),
                    collapse = ""
                )
        )
}))

ggplot(
    history_data,
    aes(x = history, y = n, fill = observed)
) +
    geom_col() +
    facet_wrap(~scenario, scales = "free_y") +
    scale_fill_manual(
        values = c(`FALSE` = "#d95f02", `TRUE` = "#1b9e77"),
        labels = c(`FALSE` = "All-zero / hidden", `TRUE` = "Observed"),
        name = NULL
    ) +
    labs(
        title = "Capture-history frequencies",
        x = "Capture history",
        y = "Individuals"
    )
```

The history table is also a mechanism check. Under conditional independence,
the observed frequencies should reflect the list probabilities and the
covariate composition. A systematic excess or deficit in particular pairwise
histories would be evidence to investigate dependence or latent heterogeneity
before comparing estimators.

# Pairwise Overlap Diagnostics

The overlap summary compares each observed pairwise intersection with the
intersection expected from the two marginal list counts under independence.
The ratio `observed_to_expected` is centered around 1 when the observed overlap
agrees with that reference. Values above 1 indicate excess overlap, while
values below 1 indicate less overlap than expected.

```{r overlap-heatmap}
#| fig-width: 16
#| fig-height: 7
overlap_data <- bind_rows(lapply(scenario_summaries, function(summary) {
    summary$overlaps |>
        filter(
            scope %in% c("full", "observed"),
            !is.na(observed_to_expected)
        ) |>
        mutate(scenario = summary$metadata$scenario_name)
}))

overlap_heatmap_data <- bind_rows(
    overlap_data,
    overlap_data |>
        transmute(
            scope,
            capture_1 = capture_2,
            capture_2 = capture_1,
            observed_to_expected,
            scenario
        )
) |>
    distinct(scope, capture_1, capture_2, scenario, .keep_all = TRUE)

ggplot(
    overlap_heatmap_data,
    aes(x = capture_1, y = capture_2, fill = observed_to_expected)
) +
    geom_tile() +
    geom_label(
        aes(label = sprintf("%.2f", observed_to_expected)),
        size = 2.8,
        linewidth = 0,
        fill = scales::alpha("white", 0.75)
    ) +
    facet_grid(scope ~ scenario) +
    coord_fixed() +
    scale_fill_gradient2(
        low = "#4575b4",
        mid = "white",
        high = "#d73027",
        midpoint = 1,
        name = "Observed / expected"
    ) +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.spacing = grid::unit(1, "lines")
    ) +
    labs(
        title = "Pairwise overlap relative to marginal independence",
        x = "Capture list",
        y = "Capture list"
    )
```

The `full` rows are the cleanest check of the generating mechanism because all
individuals are available. The `observed` rows answer a different
question: they describe dependence after conditioning on at least one capture,
so selection can create apparent association even when the lists are
conditionally independent in the full population.

The following table isolates the pair used by the direct-dependence mechanism.
The continuous latent and latent-class scenarios should also show excess
marginal overlap, because high-capture individuals are more likely to appear
in both lists. The direct-dependence scenario changes the pair conditional on
the first list and therefore represents dependence that remains after
accounting for the generated population mechanism.

```{r dependence-verification}
dependence_verification <- bind_rows(lapply(
    scenario_summaries,
    function(summary) {
        summary$overlaps |>
            filter(
                scope == "full",
                capture_1 == "capture_1",
                capture_2 == "capture_2"
            ) |>
            transmute(
                scenario = summary$metadata$scenario_name,
                observed_to_expected,
                jaccard
            )
    }
))

baseline_ratio <- dependence_verification$observed_to_expected[
    dependence_verification$scenario == "mid_capture"
]
latent_ratio <- dependence_verification$observed_to_expected[
    dependence_verification$scenario == "latent_heterogeneity"
]
pairwise_ratio <- dependence_verification$observed_to_expected[
    dependence_verification$scenario == "direct_pairwise_dependence"
]
class_ratio <- dependence_verification$observed_to_expected[
    dependence_verification$scenario == "latent_classes"
]
graph_summary <- scenario_summaries$graph_dependence$overlaps
graph_positive_ratio <- graph_summary$observed_to_expected[
    graph_summary$scope == "full" &
        graph_summary$capture_1 == "capture_1" &
        graph_summary$capture_2 == "capture_2"
]
graph_negative_ratio <- graph_summary$observed_to_expected[
    graph_summary$scope == "full" &
        graph_summary$capture_1 == "capture_2" &
        graph_summary$capture_2 == "capture_3"
]
stopifnot(
    latent_ratio > baseline_ratio,
    pairwise_ratio > baseline_ratio,
    class_ratio > baseline_ratio,
    graph_positive_ratio > 1,
    graph_negative_ratio < 1
)

dependence_verification
```

# Estimator Error Decomposition

The estimator only receives `observed_data`, so its population-size estimate
can be compared with the simulation truth without revealing the all-zero
records. For each list pair, the estimated hidden count is the estimated total
minus the observed count. Because the observed count is fixed for a simulated
dataset, total population error and hidden-component error are numerically the
same quantity:

$$
\widehat{N}_{0} = \widehat{N} - N_{\mathrm{observed}}, \qquad
\widehat{N} - N = \widehat{N}_{0} - N_{0}.
$$

```{r estimator-comparison}
estimator_comparison <- bind_rows(lapply(scenario_data, function(simulation) {
    capture_columns <- simulation$metadata$capture_columns
    estimates <- estimate_capture_prob(
        simulation$observed_data[, capture_columns, drop = FALSE],
        n_lists = length(capture_columns)
    )
    estimates |>
        mutate(
            scenario = simulation$metadata$scenario_name,
            n_total = simulation$truth$n_total,
            n_observed = simulation$truth$n_observed,
            n_unobserved = simulation$truth$n_unobserved,
            estimated_unobserved = n - n_observed,
            total_error = n - n_total,
            unobserved_error = estimated_unobserved - n_unobserved
        )
}))

stopifnot(
    all.equal(
        estimator_comparison$total_error,
        estimator_comparison$unobserved_error
    )
)

estimator_comparison
```

The pair-specific estimates make mechanism sensitivity visible. A large error
for the directly dependent pair is expected because that pair violates the
independence structure used by the estimator. Latent heterogeneity can affect
all pairs through marginal association, even though the lists are independent
conditional on the unobserved class or propensity.

```{r estimator-error-plot}
ggplot(
    estimator_comparison,
    aes(x = scenario, y = total_error, fill = listpair)
) +
    geom_col(position = position_dodge(width = 0.8)) +
    geom_hline(yintercept = 0, linewidth = 0.4) +
    labs(
        title = "Pairwise population-size error by mechanism",
        x = NULL,
        y = "Estimated total minus true total",
        fill = "List pair"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Repeated Estimator Performance

The single-simulation comparison above shows mechanism-specific errors, but a
simulation envelope needs independent replicates. These factories preserve the
same data-generating mechanisms while allowing the seed to vary by replicate.

```{r repeated-estimator-performance}
scenario_factories <- list(
    mid_capture = function(seed) {
        simulate_data(
            n_individuals = 5000,
            n_captures = 3,
            p_captures = c(0.50, 0.50, 0.50),
            seed = seed
        )
    },
    low_capture = function(seed) {
        simulate_data(
            n_individuals = 5000,
            n_captures = 3,
            p_captures = c(0.20, 0.20, 0.20),
            seed = seed
        )
    },
    imbalanced_capture = function(seed) {
        simulate_data(
            n_individuals = 5000,
            n_captures = 3,
            p_captures = c(0.80, 0.35, 0.10),
            seed = seed
        )
    },
    latent_heterogeneity = function(seed) {
        simulate_data(
            n_individuals = 5000,
            n_captures = 3,
            p_captures = rep(0.5, 3),
            latent_sd = 1,
            seed = seed
        )
    },
    direct_pairwise_dependence = function(seed) {
        simulate_data(
            n_individuals = 5000,
            n_captures = 3,
            p_captures = rep(0.5, 3),
            dependence_pair = c(1, 2),
            dependence_strength = 2,
            seed = seed
        )
    },
    latent_classes = function(seed) {
        simulate_data(
            n_individuals = 5000,
            latent_class_probabilities = c(0.75, 0.25),
            latent_class_p_captures = rbind(
                c(0.8, 0.8, 0.8),
                c(0.2, 0.2, 0.2)
            ),
            seed = seed
        )
    },
    graph_dependence = function(seed) {
        simulate_data(
            n_individuals = 5000,
            n_captures = 3,
            p_captures = rep(0.5, 3),
            dependence_edges = data.frame(
                from = c(1, 2),
                to = c(2, 3),
                strength = c(2, -2)
            ),
            seed = seed
        )
    }
)

run_estimator_replicate <- function(factory, scenario_name, replicate, seed) {
    simulation <- factory(seed)
    capture_columns <- simulation$metadata$capture_columns
    estimate_capture_prob(
        simulation$observed_data[, capture_columns, drop = FALSE],
        n_lists = length(capture_columns)
    ) |>
        mutate(
            scenario = scenario_name,
            replicate = replicate,
            seed = seed,
            n_total = simulation$truth$n_total,
            n_observed = simulation$truth$n_observed,
            n_unobserved = simulation$truth$n_unobserved,
            bias = n - n_total,
            relative_bias = bias / n_total,
            coverage = ci_l <= n_total & ci_u >= n_total,
            interval_width = ci_u - ci_l
        )
}

replicate_seeds <- 20261006 + seq_len(20)
replicate_estimates <- bind_rows(lapply(
    names(scenario_factories),
    function(scenario_name) {
        bind_rows(lapply(seq_along(replicate_seeds), function(replicate) {
            run_estimator_replicate(
                scenario_factories[[scenario_name]],
                scenario_name,
                replicate,
                replicate_seeds[[replicate]]
            )
        }))
    }
))

replicate_summary <- replicate_estimates |>
    group_by(scenario, listpair) |>
    summarise(
        replicates = n(),
        mean_bias = mean(bias),
        median_bias = median(bias),
        rmse = sqrt(mean(bias^2)),
        coverage = mean(coverage),
        mean_interval_width = mean(interval_width),
        interval_width_q05 = quantile(interval_width, 0.05),
        interval_width_q95 = quantile(interval_width, 0.95),
        .groups = "drop"
    )

stopifnot(
    nrow(replicate_estimates) ==
        length(scenario_factories) * length(replicate_seeds) * 3,
    length(unique(replicate_estimates$scenario)) == length(scenario_factories),
    all(
        c(
            "bias",
            "relative_bias",
            "coverage",
            "interval_width"
        ) %in%
            names(replicate_estimates)
    )
)

replicate_summary
```

The envelope reports signed bias, RMSE, empirical coverage, and the central
90% interval-width range for each mechanism and list pair. Coverage is computed
against the known simulated total, while the estimator still receives only the
observed union.

```{r repeated-estimator-plot}
ggplot(
    replicate_summary,
    aes(x = scenario, y = mean_bias, fill = listpair)
) +
    geom_col(position = position_dodge(width = 0.8)) +
    geom_hline(yintercept = 0, linewidth = 0.4) +
    labs(
        title = "Mean estimator bias across simulation replicates",
        x = NULL,
        y = "Mean estimated total minus true total",
        fill = "List pair"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Next Analysis Steps

This notebook now validates both the fixed-probability baseline and the
public covariate-dependent API. The next analysis additions should be:

1.  Add covariate-model misspecification and inspect capture probability
    overlap by covariate value.
2.  Compare estimator methods under covariate-model misspecification.