---
title: "Simulation Results Visualization"
author:
    - name: "Matthew Carroll"
---

```{r}
library(dplyr)
library(purrr)
library(ggplot2)
```

## Load Simulation Results

```{r}
result_files <- list.files(
    file.path("data", "intercepts_and_regress_coefs"),
    pattern = "^final_results_[-.0-9]+int_[-.0-9]+rc\\.csv$",
    full.names = TRUE
)
```

## Combine Results

```{r}

simulation_results <- map_dfr(result_files, function(file) {
    scenario <- sub("^final_results_", "", basename(file))
    scenario <- sub("\\.csv$", "", scenario)
    parameters <- strcapture(
        "^([-.0-9]+)int_([-.0-9]+)rc$",
        scenario,
        proto = list(intercept = numeric(), regression_coef = numeric())
    )

    read.csv(file) |>
        mutate(
            intercept = parameters$intercept,
            regression_coef = parameters$regression_coef
        )
}) |>
    mutate(
        method = factor(method, levels = unique(method)),
        intercept_label = factor(
            sprintf("%.1f", intercept),
            levels = sprintf("%.1f", sort(unique(intercept)))
        ),
        regression_coef_label = factor(
            sprintf("%.1f", regression_coef),
            levels = sprintf("%.1f", sort(unique(regression_coef)))
        )
    )
```

## Coverage

Coverage is the proportion of simulation confidence intervals that contain the
true relative bias of zero. The dashed contour marks nominal 95% coverage.

```{r}
coverage_plot <- ggplot(
    simulation_results,
    aes(x = intercept_label, y = method, fill = coverage)
) +
    geom_tile(color = "white", linewidth = 0.6) +
    geom_text(aes(label = sprintf("%.0f%%", coverage * 100)), size = 3.2) +
    facet_wrap(~regression_coef_label, nrow = 1) +
    scale_fill_gradientn(
        colors = c("#b2182b", "#fddbc7", "#d1e5f0", "#2166ac"),
        limits = c(0, 1),
        name = "Coverage"
    ) +
    labs(
        title = "Confidence Interval Coverage by Simulation Scenario",
        x = "Capture-probability intercept",
        y = "Estimation method",
        subtitle = "Facets show the covariate regression coefficient"
    ) +
    theme_minimal(base_size = 14) +
    theme(
        panel.grid = element_blank(),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 14),
        strip.text = element_text(size = 14, face = "bold"),
        plot.title = element_text(size = 17, face = "bold"),
        plot.subtitle = element_text(size = 13),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 13)
    )

coverage_plot
```

## Mean Estimates and Intervals

Each point is a method's mean relative-bias estimate and each line spans the
mean lower and upper confidence-interval endpoints across simulations. Panels
use independent vertical scales so estimates near zero remain readable.

```{r mean_estimates}
interval_plot <- ggplot(
    simulation_results,
    aes(x = intercept_label, y = estimate, color = method)
) +
    geom_hline(yintercept = 0, color = "#003771", linetype = "dashed") +
    geom_errorbar(
        aes(ymin = lower_ci, ymax = upper_ci),
        width = 0.18,
        linewidth = 0.5
    ) +
    geom_point(size = 2) +
    facet_grid(
        regression_coef_label ~ method,
        scales = "free_y",
        labeller = label_both
    ) +
    labs(
        title = "Mean Relative-Bias Estimates and Confidence Intervals",
        x = "Capture-probability intercept",
        y = "Mean relative bias",
        caption = "The dashed horizontal line is the true relative bias (0)."
    ) +
    scale_color_manual(
        values = c(
            "AIC-Negbin" = "#003771",
            "AIC-Poisson" = "#3d9be9",
            "DR" = "#f0325f",
            "PI" = "#1565a8",
            "Stepwise-Negbin" = "#d51f49",
            "Stepwise-Poisson" = "#6bb5ed",
            "TMLE" = "#001f42"
        )
    ) +
    theme_minimal(
        base_size = 10,
        base_family = "IBM Plex Sans"
    ) +
    theme(
        plot.background = element_rect(fill = "#ffffff", color = NA),
        panel.background = element_rect(fill = "#ffffff", color = NA),
        legend.position = "none",
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title = element_text(family = "Helvetica"),
        plot.title = element_text(
            family = "Helvetica",
            face = "bold",
            size = 14
        ),
        strip.text = element_text(
            family = "Helvetica",
            face = "bold",
            size = 8
        )
    )

interval_plot
```

## Removed AIC-Poisson

Because AIC-Poisson produces the largest CIs, we remove it from the plot to better visualize the other methods. The dashed contour marks nominal 95% coverage.

```{r mean_estimates_no_aicp}
interval_plot <- simulation_results |>
    filter(method != "AIC-Poisson") |>
    ggplot(
        aes(x = intercept_label, y = estimate, color = method)
    ) +
    geom_hline(yintercept = 0, color = "#003771", linetype = "dashed") +
    geom_errorbar(
        aes(ymin = lower_ci, ymax = upper_ci),
        width = 0.18,
        linewidth = 0.5
    ) +
    geom_point(size = 2) +
    facet_grid(
        regression_coef_label ~ method,
        scales = "free_y",
        labeller = label_both
    ) +
    labs(
        title = "Mean Relative-Bias Estimates and Confidence Intervals",
        x = "Capture-probability intercept",
        y = "Mean relative bias",
        caption = "The dashed horizontal line is the true relative bias (0)."
    ) +
    scale_color_manual(
        values = c(
            "AIC-Negbin" = "#003771",
            "DR" = "#f0325f",
            "PI" = "#1565a8",
            "Stepwise-Negbin" = "#d51f49",
            "Stepwise-Poisson" = "#6bb5ed",
            "TMLE" = "#001f42"
        )
    ) +
    theme_minimal(
        base_size = 10,
        base_family = "IBM Plex Sans"
    ) +
    theme(
        plot.background = element_rect(fill = "#ffffff", color = NA),
        panel.background = element_rect(fill = "#ffffff", color = NA),
        legend.position = "none",
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title = element_text(family = "Helvetica"),
        plot.title = element_text(
            family = "Helvetica",
            face = "bold",
            size = 14
        ),
        strip.text = element_text(
            family = "Helvetica",
            face = "bold",
            size = 8
        )
    )

interval_plot
```